\textbf{OptiClust: Improved method for assigning amplicon-based sequence data to operational taxonomic units}

### 2017-02-03 submission/Westcott_OptiClust_mSphere_2016.tex
Re render with edits

\documentclass[11pt,]{article}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \ifxetex
    \usepackage{mathspec}
  \else
    \usepackage{fontspec}
  \fi
  \defaultfontfeatures{Ligatures=TeX,Scale=MatchLowercase}
  \newcommand{\euro}{€}
\fi
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
% use microtype if available
\IfFileExists{microtype.sty}{%
\usepackage{microtype}
\UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\usepackage[margin=1.0in]{geometry}
\usepackage{hyperref}
\PassOptionsToPackage{usenames,dvipsnames}{color} % color is loaded by hyperref
\hypersetup{unicode=true,
            pdftitle={OptiClust: Improved method for assigning amplicon-based sequence data to operational taxonomic units},
            pdfborder={0 0 0},
            breaklinks=true}
\urlstyle{same}  % don't use monospace font for urls
\usepackage{longtable,booktabs}
\usepackage{graphicx,grffile}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt plus 2pt minus 1pt}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{0}

@@

A growing number of algorithms have been developed to cluster sequences
into OTUs. These algorithms can be classified into three general
categories. The first category of algorithms has been termed
closed-reference or phylotyping (13, 14). Sequences are compared to a
reference collection and clustered based on the reference sequences that
they are similar to. This approach is fast; however, the method
struggles when a sequence is similar to multiple reference sequences
that may have different taxonomies and when it is not similar to
sequences in the reference. The second category of algorithms has been
called \emph{de novo} because they assign sequences to OTUs without the
use of a reference (14). These include hierarchical algorithms such as
nearest, furthest, and average neighbor (15) and algorithms that employ
heuristics such as abundance or distance-based greedy clustering as
implemented in USEARCH (16) or VSEARCH (17), Sumaclust, OTUCLUST (18),
and Swarm (19). \emph{De novo} methods are agglomerative and tend to be
more computationally intense. It has proven difficult to know which
method generates the best assignments. A third category of algorithm is
open reference clustering, which is a hybrid approach (3, 14). Here
sequences are assigned to OTUs using closed-reference clustering and
sequences that are not within a threshold of a reference sequence are
then clustered using a \emph{de novo} approach. This category blends the
strengths and weaknesses of the other method and adds the complication
that closed-reference and \emph{de novo} clustering use different OTU
definitions. These algorithms take different approaches to handling
large datasets to minimize the time and memory requirements while
attempting to assign sequences to meaningful OTUs.

Several metrics have emerged for assessing the quality of OTU assignment
algorithms. These have included the time and memory required to run the
algorithm (3, 19--21), agreement between OTU assignments and the
sequences' taxonomy (19, 21--31), sensitivity of an algorithm to
stochastic processes (32), the number of OTUs generated by the algorithm
(22, 33), and the ability to regenerate the assignments made by other
algorithms (3, 34). Unfortunately, these methods fail to directly
quantify the quality of the OTU assignments. An algorithm may complete
with minimal time and memory requirements or generate an idealized
number of OTUs, but the composition of the OTUs could be incorrect.
These metrics also tend to be subjective. For instance, a method may
appear to recapitulate the taxonomy of a synthetic community with known
taxonomic structure, but do a poor job when applied to real communities
with poorly defined taxonomic structure or for sequences that are prone
to misclassification. As an alternative, we developed an approach to
objectively benchmark the clustering quality of OTU assignments (13, 35,
36). This approach counts the number of true positives (TP), true
negatives (TN), false positives (FP), and false negatives (FN) based on
the pairwise distances. Sequence pairs that are within the
user-specified threshold and are clustered together represent TPs and
those in different OTUs are FNs. Those sequence pairs that have a
distance larger than the threshold and are not clustered in the same OTU
are TNs and those in the same OTU are FPs. These values can be
synthesized into a single correlation coefficient, the Matthews'
Correlation Coefficient (MCC), which measures the correlation between
observed and predicted classifications and is robust to cases where
there is an uneven distribution across the confusion matrix (37).
Consistently, the average neighbor algorithm was identified as among the
best or the best algorithm. Other hierarchical algorithms such as
furthest and nearest neighbor, which seek to minimize the number of FP
and FN, respectively, fared significantly worse. The distance-based
greedy clustering as implemented in VSEARCH has also performed well. The
computational resources required to complete the average neighbor
algorithm can be significant for large datasets and so there is a need
for an algorithm that efficiently produces consistently high quality OTU
assignments.

These previous efforts have assessed the quality of the clusters after
the completion of the algorithm. In the current study we developed and
benchmarked a new \emph{de novo} clustering algorithm that uses real
time calculation of the MCC to direct the progress of the clustering.
The result is the OptiClust algorithm, which produces significantly
better sequence assignments while making efficient use of computational
resources.

@@

\textbf{\emph{OptiClust algorithm.}} The OptiClust algorithm uses the
pairs of sequences that are within a desired threshold of each other
(e.g.~0.03), a list of all sequence names in the dataset, and the metric
that should be used to assess clustering quality. A detailed description
of the algorithm is provided for a toy dataset in the Supplementary
Material. Briefly, the algorithm starts by placing each sequence either
within its own OTU or into a single OTU. The algorithm proceeds by
interrogating each sequence and re-calculating the metric for the cases
where the sequence stays in its current OTU, is moved to each of the
other OTUs, or is moved into a new OTU. The location that results in the
best clustering quality indicates whether the sequence should remain in
its current OTU or be moved to a different or new OTU. Each iteration
consists of interrogating every sequence in the dataset. Although
numerous options are available for optimizing the clusters and for
assessing the quality of the clusters within the mothur-based
implementation of the algorithm (e.g.~sensitivity, specificity,
accuracy, F1 score, etc.), the default metric for optimization and
assessment is MCC because it includes all four parameters from the
confusion matrix (Figure S1). The algorithm continues until the
optimization metric stabilizes or until it reaches a defined stopping
criteria (Figure S2).

\textbf{\emph{OptiClust-generated OTUs are more robust than those from
other methods.}} To evaluate the OptiClust algorithm and compare its
performance to other algorithms, we utilized six datasets including two
synthetic communities and four previously published large datasets
generated from soil, marine, human, and murine samples (Table 1). When
we seeded the OptiClust algorithm with each sequence in a separate OTU
and ran the algorithm until complete convergence, the MCC values
averaged 15.2 and 16.5\% higher than the OTUs using average neighbor and
distance-based greedy clustering (DGC) with VSEARCH, respectively
(Figure 1). The number of OTUs formed by the various methods was
negatively correlated with their MCC value (\(\rho\)=-0.47; p=0). The
OptiClust algorithm was considerably faster than the hierarchical
algorithms and somewhat slower than the heuristic-based algorithms.
Across the six datasets, the OptiClust algorithm was 94.6-times faster
than average neighbor and just as fast as DGC with VSEARCH. The human
dataset was a challenge for a number of the algorithms. OTUCLUST and
SumaClust were unable to cluster the human dataset in less than 50 hours
and the average neighbor algorithm required more than 45 GB of RAM. The
USEARCH-based methods were unable to cluster the human data using the
32-bit free version of the software that limits the amount of RAM to
approximately 3.5 GB. These data demonstrate that OptiClust generated
significantly more robust OTU assignments than existing methods across a
diverse collection of datasets with performance that was comparable to
popular methods.

\textbf{\emph{OptiClust stopping criteria.}} By default, the
mothur-based implementation of the algorithm stops when the optimization
metric changes by less than 0.0001; however, this can be altered by the
user. This implementation also allows the user to stop the algorithm if
a maximum number of iterations is exceeded. By default mothur uses a
maximum value of 100 iterations. The justification for allowing
incomplete convergence was based on the observation that numerous
iterations are performed that extend the time required to complete the
clustering with minimal improvement in clustering. We evaluated the
results of clustering to partial convergence (i.e.~a change in the MCC
value that was less than 0.0001) or until complete convergence of the
MCC value (i.e.~until it did not change between iterations) when seeding
the algorithm with each sequence in a separate OTU (Figure 1). The small
difference in MCC values between the output from partial and complete
convergence resulted in a difference in the median number of OTUs that
ranged between 1.5 and 17.0 OTUs. This represented a difference of less
than 0.15\%. Among the four natural datasets, between 3 and 6 were
needed to achieve partial convergence and between 8 and 12.50 iterations
were needed to reach full convergence. The additional steps required
between 1.4 and 1.7 times longer to complete the algorithm. These
results suggest that achieving full convergence of the optimization
metric adds computational effort; however, considering full convergence
took between 2 and 17 minutes the extra effort was relatively small.
Although the mothur's default setting is partial convergence, the
remainder of our analysis used complete convergence to be more
conservative.

@@

\textbf{\emph{OptiClust-generated OTUs are as stable as those from other
algorithms.}} One concern that many have with \emph{de novo} clustering
algorithms is that their output is sensitive to the initial order of the
sequences because each algorithm must break ties where a sequence could
be assigned to multiple OTUs. An additional concern with the OptiClust
algorithm is that it may stabilize at a local optimum. To evaluate these
concerns we compared the results obtained using ten randomizations of
the order that sequences were given to the algorithm. The median the
coefficient of variation across the six datasets for MCC values obtained
from the replicate clusterings using OptiClust was 0.1\% (Figure 1). We
also measured the coefficient of variation for the number of OTUs across
the six datasets for each method. The median coefficient of variation
for the number of OTUs generated using OptiClust was 0.1\%. Confirming
our previous results, all of the methods we tested were stable to
stochastic processes. Of the methods that involved randomization, the
coefficient of variation for MCC values was considerably smaller with
OptiClust than the other methods and the coefficient of variation for
the number of OTUs was comparable to the other methods. The variation
observed in clustering quality suggested that the algorithm does not
appear to converge to a locally optimum MCC value. More importantly, the
random variation does yield output of a similarly high quality.

\textbf{\emph{Time and memory required to complete Optimization-based
clustering scales efficiently.}} Although not as important as the
quality of clustering, the amount of time and memory required to assign
sequences to OTUs is a legitimate concern. We observed that the time
required to complete the OptiClust algorithm (Figure 1C) paralleled the
number of pairwise distances that were smaller than 0.03. To further
evaluate how the speed and memory usage scaled with the number of
sequences in the dataset, we measured the time required and maximum RAM
usage to cluster 20, 40, 60, 80, and 100\% of the unique sequences from
each of the natural datasets using the OptiClust algorithm (Figure 2).
Within each iteration of the algorithm, each sequence is compared to
every other sequence and each comparison requires a recalculation of the
confusion matrix. This would result in a worst case algorithmic
complexity on the order of N\^{}3, where N is the number of unique
sequences. Because the algorithm only needs to keep track of the
sequence pairs that are within the threshold of each other, it is likely
that the implementation of the algorithm is more efficient. To
empirically determine the algorithmic complexity, we fit a power law
function to the data in Figure 2A. We observed power coefficients
between 1.7 and 2.5 for the marine and human datasets, respectively. The
algorithm requires storing a matrix that contains the pairs of sequences
that are close to each other as well as a matrix that indicates which
sequences are clustered together. The memory required to store these
matrices is on the order of N\^{}2, where N is the number of unique
sequences. In fact, when we fit a power law function to the data in
Figure 2B, the power coefficients were 1.9. Using the four natural
community datasets, doubling the number of sequences in a dataset would
increase the time required to cluster the data by 4 to 8-fold and
increase the RAM required by 4-fold. It is possible that future
improvements to the implementation of the algorithm could improve this
performance.

\textbf{\emph{Cluster splitting heuristic generates OTUs that are as
good as non-split approach.}} We previously described a heuristic to
accelerate OTU assignments where sequences were first classified to
taxonomic groups and within each taxon sequences were assigned to OTUs
using the average neighbor clustering algorithm (13). This method is
similar to open reference clustering except that in our approach all
sequences are subjected to \emph{de novo} clustering following
classification whereas in open reference clustering only those sequences
that cannot be classified are subjected to \emph{de novo} clustering.
Our cluster splitting approach accelerated the clustering and reduced
the memory requirements because the number of unique sequences is
effectively reduced by splitting sequences across taxonomic groups.
Furthermore, because sequences in different taxonomic groups are assumed
to belong to different OTUs they are independent, which permits
parallelization and additional reduction in computation time. Reduction
in clustering quality are encountered in this approach if there are
errors in classification or if two sequences within the desired
threshold belong to different taxonomic groups. It is expected that
these errors would increase as the taxonomic level goes from kingdom to
genus. To characterize the clustering quality, we classified each
sequence at each taxonomic level and calculated the MCC values using
OptiClust, average neighbor, and DGC with VSEARCH when splitting at each
taxonomic level (Figure 3). For each method, the MCC values decreased as
the taxonomic resolution increased; however, the decrease in MCC as not
as large as the difference between clustering methods. As the resolution
of the taxonomic levels increased, the clustering quality remained high,
relative to clusters formed from the entire dataset
(i.e.~kingdom-level). The MCC values when splitting the datasets at the
class and genus levels were within 98.0 and 93.0\%, respectively, of the
MCC values obtained from the entire dataset. These decreases in MCC
value resulted in the formation of as many as 4.7 and 22.5\% more OTUs,
respectively, than were observed from the entire dataset. These errors
were due to the generation of additional false negatives due to
splitting similar sequences into different taxonomic groups. For the
datasets included in the current analysis, the use of the cluster
splitting heuristic was probably not worth the loss in clustering
quality. However, as datasets become larger, it may be necessary to use
the heuristic to clustering the data into OTUs.

@@

Myriad methods have been proposed for assigning 16S rRNA gene sequences
to OTUs that each claim improved performance based on speed, memory
usage, representation of taxonomic information, and number of OTUs. Each
of these metrics is subjective and do not actually indicate the quality
of the clustering. This led us to propose using the MCC as a metric for
assessing the quality of clustering, post hoc. Here, we described a new
clustering method that seeks to optimize clustering based on an
objective criterion that measures clustering quality in real time. In
the OptiClust algorithm clustering is driven by optimizing a metric that
assesses whether any two sequences should be grouped into the same OTU.
The result is clusters that are significantly more robust and is
efficient in the time and memory required to cluster the sequences into
OTUs. This makes it more tractable to analyze large datasets without
sacrificing clustering quality as was previously necessary using
heuristic methods.

The cluster optimization procedure is dependent on the metric that is
chosen for optimization. We employed the MCC because it includes the
four values from a confusion matrix. Other algorithms such as the
furthest neighbor and nearest neighbor algorithms minimize the number of
FP and FN, respectively; however, these suffer because the number of FN
and FP are not controlled, respectively (13, 15). Alternatively, one
could optimize based on the sensitivity, specificity, or accuracy, which
are each based on two values from the confusion matrix or they could
optimize based on the F1 score, which is based on three values from the
confusion matrix. Because these metrics do not balance all four
parameters equally, it is likely that one parameter will dominate in the
optimization procedure. For example, optimizing for sensitivity could
lead to a large number of FPs. More FPs increases the number of OTUs
while more FNs collapses OTUs together. It is difficult to know which is
worse since community richness and diversity are linked to the number of
OTUs. In addition, increasing the number of FNs would overstate the
differences between communities while increasing the number of FPs would
overstate their similarity. Therefore, it is important to jointly
minimize the number of FPs and FNs. With this in mind, we decided to
optimize utilizing the MCC. It is possible that other metrics that
balance the four parameters could be developed and employed for
optimization of the clustering.

@@

\textbf{\emph{Sequence data and processing steps.}} To evaluate the
OptiClust and the other algorithms we created two synthetic sequence
collections and four sequence collections generated from previously
published studies. The V4 region of the 16S rRNA gene was used from all
datasets because it is a popular region that can be fully sequenced with
two-fold coverage using the commonly used MiSeq sequencer from Illumina
(7). The method for generating the simulated datasets followed the
approach used by Kopylova et al. (33) and Schloss (35). Briefly, we
randomly selected 10,000 uniques V4 fragments from 16S rRNA gene
sequences that were unique from the SILVA non-redundant database (38). A
community with an even relative abundance profile was generated by
specifying that each sequence had a frequency of 100 reads. A community
with a staggered relative abundance profile was generated by specifying
that the abundance of each sequence was a randomly drawn integer sampled
from a uniform distribution between 1 and 200. Sequence collections
collected from human feces (39), murine feces (40), soil (41), and
seawater (42) were used to characterize the algorithms' performance with
natural communities. These sequence collections were all generated using
paired 150 or 250 nt reads of the V4 region. We re-processed all of the
reads using a common analysis pipeline that included quality score-based
error correction (7), alignment against a SILVA reference database (38,
43), screening for chimeras using UCHIME (9), and classification using a
naive Bayesian classifier with the RDP training set requiring an 80\%
confidence score (10).

>>>>

\documentclass[11pt,]{article}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \ifxetex
    \usepackage{mathspec}
    \usepackage{xltxtra,xunicode}
  \else
    \usepackage{fontspec}
  \fi
  \defaultfontfeatures{Mapping=tex-text,Scale=MatchLowercase}
  \newcommand{\euro}{€}
\fi
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
% use microtype if available
\IfFileExists{microtype.sty}{%
\usepackage{microtype}
\UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\usepackage[margin=1.0in]{geometry}
\ifxetex
  \usepackage[setpagesize=false, % page size defined by xetex
              unicode=false, % unicode breaks when used with xetex
              xetex]{hyperref}
\else
  \usepackage[unicode=true]{hyperref}
\fi
\hypersetup{breaklinks=true,
            bookmarks=true,
            pdfauthor={},
            pdftitle={OptiClust: Improved method for assigning amplicon-based sequence data to operational taxonomic units},
            colorlinks=true,
            citecolor=blue,
            urlcolor=blue,
            linkcolor=magenta,
            pdfborder={0 0 0}}
\urlstyle{same}  % don't use monospace font for urls
\usepackage{longtable,booktabs}
\usepackage{graphicx,grffile}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt plus 2pt minus 1pt}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{0}

@@

A growing number of algorithms have been developed to cluster sequences
into OTUs. These algorithms can be classified into three general
categories. The first category of algorithms has been termed
closed-reference or phylotyping (13, 14). Sequences are compared to a
reference collection and clustered based on the reference sequences that
they are similar to. This approach is fast; however, the method
struggles when a sequence is similar to multiple reference sequences
that may have different taxonomies and when it is not similar to
sequences in the reference (15). The second category of algorithms has
been called \emph{de novo} because they assign sequences to OTUs without
the use of a reference (14). These include hierarchical algorithms such
as nearest, furthest, and average neighbor (16) and algorithms that
employ heuristics such as abundance or distance-based greedy clustering
as implemented in USEARCH (17) or VSEARCH (18), Sumaclust, OTUCLUST
(19), and Swarm (20). \emph{De novo} methods are agglomerative and tend
to be more computationally intense. It has proven difficult to know
which method generates the best assignments. A third category of
algorithm is open reference clustering, which is a hybrid approach (3,
14). Here sequences are assigned to OTUs using closed-reference
clustering and sequences that are not within a threshold of a reference
sequence are then clustered using a \emph{de novo} approach. This
category blends the strengths and weaknesses of the other method and
adds the complication that closed-reference and \emph{de novo}
clustering use different OTU definitions. These three categories of
algorithms take different approaches to handling large datasets to
minimize the time and memory requirements while attempting to assign
sequences to meaningful OTUs.

Several metrics have emerged for assessing the quality of OTU assignment
algorithms. These have included the time and memory required to run the
algorithm (3, 20--22), agreement between OTU assignments and the
sequences' taxonomy (20, 22--32), sensitivity of an algorithm to
stochastic processes (33), the number of OTUs generated by the algorithm
(23, 34), and the ability to regenerate the assignments made by other
algorithms (3, 35). Unfortunately, these methods fail to directly
quantify the quality of the OTU assignments. An algorithm may complete
with minimal time and memory requirements or generate an idealized
number of OTUs, but the composition of the OTUs could be incorrect.
These metrics also tend to be subjective. For instance, a method may
appear to recapitulate the taxonomy of a synthetic community with known
taxonomic structure, but do a poor job when applied to real communities
with poorly defined taxonomic structure or for sequences that are prone
to misclassification. As an alternative, we developed an approach to
objectively benchmark the clustering quality of OTU assignments (13, 15,
36). This approach counts the number of true positives (TP), true
negatives (TN), false positives (FP), and false negatives (FN) based on
the pairwise distances. Sequence pairs that are within the
user-specified threshold and are clustered together represent TPs and
those in different OTUs are FNs. Those sequence pairs that have a
distance larger than the threshold and are not clustered in the same OTU
are TNs and those in the same OTU are FPs. These values can be
synthesized into a single correlation coefficient, the Matthews
correlation coefficient (MCC), which measures the correlation between
observed and predicted classifications and is robust to cases where
there is an uneven distribution across the confusion matrix (37).
Consistently, the average neighbor algorithm was identified as among the
best or the best algorithm. Other hierarchical algorithms such as
furthest and nearest neighbor, which do not permit the formation of FPs
or FNs, respectively, fared significantly worse. The distance-based
greedy clustering as implemented in VSEARCH has also performed well. The
computational resources required to complete the average neighbor
algorithm can be significant for large datasets and so there is a need
for an algorithm that efficiently produces consistently high quality OTU
assignments.

These benchmarking efforts have assessed the quality of the clusters
after the completion of the algorithm. In the current study we developed
and benchmarked a new \emph{de novo} clustering algorithm that uses real
time calculation of the MCC to direct the progress of the clustering.
The result is the OptiClust algorithm, which produces significantly
better sequence assignments while making efficient use of computational
resources.

@@

\textbf{\emph{OptiClust algorithm.}} The OptiClust algorithm uses the
pairs of sequences that are within a desired threshold of each other
(e.g.~0.03), a list of all sequence names in the dataset, and the metric
that should be used to assess clustering quality. A detailed description
of the algorithm is provided for a toy dataset in the Supplementary
Material. Briefly, the algorithm starts by placing each sequence either
within its own OTU or into a single OTU. The algorithm proceeds by
interrogating each sequence and re-calculating the metric for the cases
where the sequence stays in its current OTU, is moved to each of the
other OTUs, or is moved into a new OTU. The location that results in the
best clustering quality indicates whether the sequence should remain in
its current OTU or be moved to a different or new OTU. Each iteration
consists of interrogating every sequence in the dataset. Although
numerous options are available for optimizing the clusters and for
assessing the quality of the clusters within the mothur-based
implementation of the algorithm (e.g.~sensitivity, specificity,
accuracy, F1-score, etc.), the default metric for optimization and
assessment is MCC because it includes all four parameters from the
confusion matrix (Figure S1). The algorithm continues until the
optimization metric stabilizes or until it reaches a defined stopping
criteria.

\textbf{\emph{OptiClust-generated OTUs are more robust than those from
other methods.}} To evaluate the OptiClust algorithm and compare its
performance to other algorithms, we utilized six datasets including two
synthetic communities and four previously published large datasets
generated from soil, marine, human, and murine samples (Table 1). When
we seeded the OptiClust algorithm with each sequence in a separate OTU
and ran the algorithm until complete convergence, the MCC values
averaged 15.2 and 16.5\% higher than the OTUs using average neighbor and
distance-based greedy clustering (DGC) with VSEARCH, respectively
(Figure 1). The number of OTUs formed by the various methods was
negatively correlated with their MCC value (\(\rho\)=-0.47;
p\textless{}0.001). The OptiClust algorithm was considerably faster than
the hierarchical algorithms and somewhat slower than the heuristic-based
algorithms. Across the six datasets, the OptiClust algorithm was
94.6-times faster than average neighbor and just as fast as DGC with
VSEARCH. The human dataset was a challenge for a number of the
algorithms. OTUCLUST and SumaClust were unable to cluster the human
dataset in less than 50 hours and the average neighbor algorithm
required more than 45 GB of RAM. The USEARCH-based methods were unable
to cluster the human data using the 32-bit free version of the software
that limits the amount of RAM to approximately 3.5 GB. These data
demonstrate that OptiClust generated significantly more robust OTU
assignments than existing methods across a diverse collection of
datasets with performance that was comparable to popular methods.

\textbf{\emph{OptiClust stopping criteria.}} By default, the
mothur-based implementation of the algorithm stops when the optimization
metric changes by less than 0.0001; however, this can be altered by the
user. This implementation also allows the user to stop the algorithm if
a maximum number of iterations is exceeded. By default mothur uses a
maximum value of 100 iterations. The justification for allowing
incomplete convergence was based on the observation that numerous
iterations are performed that extend the time required to complete the
clustering with minimal improvement in clustering (Figure S2). We
evaluated the results of clustering to partial convergence (i.e.~a
change in the MCC value that was less than 0.0001) or until complete
convergence of the MCC value (i.e.~until it did not change between
iterations) when seeding the algorithm with each sequence in a separate
OTU (Figure 1). The small difference in MCC values between the output
from partial and complete convergence resulted in a difference in the
median number of OTUs that ranged between 1.5 and 17.0 OTUs. This
represented a difference of less than 0.15\%. Among the four natural
datasets, between 3 and 6 were needed to achieve partial convergence and
between 8 and 12 iterations were needed to reach full convergence. The
additional steps required between 1.4 and 1.7 times longer to complete
the algorithm. These results suggest that achieving full convergence of
the optimization metric adds computational effort; however, considering
full convergence took between 2 and 17 minutes the extra effort was
relatively small. Although the mothur's default setting is partial
convergence, the remainder of our analysis used complete convergence to
be more conservative.

@@

\textbf{\emph{OptiClust-generated OTUs are as stable as those from other
algorithms.}} One concern that many have with \emph{de novo} clustering
algorithms is that their output is sensitive to the initial order of the
sequences because each algorithm must break ties where a sequence could
be assigned to multiple OTUs. An additional concern specific to the
OptiClust algorithm is that it may stabilize at a local optimum. To
evaluate these concerns we compared the results obtained using ten
randomizations of the order that sequences were given to the algorithm.
The median coefficient of variation across the six datasets for MCC
values obtained from the replicate clusterings using OptiClust was 0.1\%
(Figure 1). We also measured the coefficient of variation for the number
of OTUs across the six datasets for each method. The median coefficient
of variation for the number of OTUs generated using OptiClust was 0.1\%.
Confirming our previous results (15), all of the methods we tested were
stable to stochastic processes. Of the methods that involved
randomization, the coefficient of variation for MCC values was
considerably smaller with OptiClust than the other methods and the
coefficient of variation for the number of OTUs was comparable to the
other methods. The variation observed in clustering quality suggested
that the algorithm does not appear to converge to a locally optimum MCC
value. More importantly, the random variation does yield output of a
similarly high quality.

\textbf{\emph{Time and memory required to complete Optimization-based
clustering scales efficiently.}} Although not as important as the
quality of clustering, the amount of time and memory required to assign
sequences to OTUs is a legitimate concern. We observed that the time
required to complete the OptiClust algorithm (Figure 1C) paralleled the
number of pairwise distances that were smaller than 0.03 (Table 1). To
further evaluate how the speed and memory usage scaled with the number
of sequences in the dataset, we measured the time required and maximum
RAM usage to cluster 20, 40, 60, 80, and 100\% of the unique sequences
from each of the natural datasets using the OptiClust algorithm (Figure
2). Within each iteration of the algorithm, each sequence is compared to
every other sequence and each comparison requires a recalculation of the
confusion matrix. This would result in a worst case algorithmic
complexity on the order of N\textsuperscript{3}, where N is the number
of unique sequences. Because the algorithm only needs to keep track of
the sequence pairs that are within the threshold of each other, it is
likely that the implementation of the algorithm is more efficient. To
empirically determine the algorithmic complexity, we fit a power law
function to the data in Figure 2A. We observed power coefficients
between 1.7 and 2.5 for the marine and human datasets, respectively. The
algorithm requires storing a matrix that contains the pairs of sequences
that are close to each other as well as a matrix that indicates which
sequences are clustered together. The memory required to store these
matrices is on the order of N\textsuperscript{2}, where N is the number
of unique sequences. In fact, when we fit a power law function to the
data in Figure 2B, the power coefficients were 1.9. Using the four
natural community datasets, doubling the number of sequences in a
dataset would increase the time required to cluster the data by 4 to
8-fold and increase the RAM required by 4-fold. It is possible that
future improvements to the implementation of the algorithm could improve
this performance.

\textbf{\emph{Cluster splitting heuristic generates OTUs that are as
good as non-split approach.}} We previously described a heuristic to
accelerate OTU assignments where sequences were first classified to
taxonomic groups and within each taxon sequences were assigned to OTUs
using the average neighbor clustering algorithm (13). This method is
similar to open reference clustering except that in our approach all
sequences are subjected to \emph{de novo} clustering following
classification whereas in open reference clustering only those sequences
that cannot be classified are subjected to \emph{de novo} clustering.
Our cluster splitting approach accelerated the clustering and reduced
the memory requirements because the number of unique sequences was
effectively reduced by splitting sequences across taxonomic groups.
Furthermore, because sequences in different taxonomic groups are assumed
to belong to different OTUs they are independent, which permits
parallelization and additional reduction in computation time. Reduction
in clustering quality is encountered in this approach if there are
errors in classification or if two sequences within the desired
threshold belong to different taxonomic groups. It is expected that
these errors would increase as the taxonomic level goes from kingdom to
genus. To characterize the clustering quality, we classified each
sequence at each taxonomic level and calculated the MCC values using
OptiClust, average neighbor, and DGC with VSEARCH when splitting at each
taxonomic level (Figure 3). For each method, the MCC values decreased as
the taxonomic resolution increased; however, the decrease in MCC was not
as large as the difference between clustering methods. As the resolution
of the taxonomic levels increased, the clustering quality remained high,
relative to clusters formed from the entire dataset
(i.e.~kingdom-level). The MCC values when splitting the datasets at the
class and genus levels were within 98.0 and 93.0\%, respectively, of the
MCC values obtained from the entire dataset. These decreases in MCC
value resulted in the formation of as many as 4.7 and 22.5\% more OTUs,
respectively, than were observed from the entire dataset. These errors
were due to the generation of additional false negatives due to
splitting similar sequences into different taxonomic groups. For the
datasets included in the current analysis, the use of the cluster
splitting heuristic was probably not worth the loss in clustering
quality. However, as datasets become larger, it may be necessary to use
the heuristic to clustering the data into OTUs.

@@

Myriad methods have been proposed for assigning 16S rRNA gene sequences
to OTUs. Each claim improved performance based on speed, memory usage,
representation of taxonomic information, and number of OTUs. Each of
these metrics is subjective and do not actually indicate the quality of
the clustering. This led us to propose using the MCC as a metric for
assessing the quality of clustering, post hoc. Here, we described a new
clustering method that seeks to optimize clustering based on an
objective criterion that measures clustering quality in real time. In
the OptiClust algorithm, clustering is driven by optimizing a metric
that assesses whether any two sequences should be grouped into the same
OTU. The result is clusters that are significantly more robust and is
efficient in the time and memory required to cluster the sequences into
OTUs. This makes it more tractable to analyze large datasets without
sacrificing clustering quality as was previously necessary using
heuristic methods.

The cluster optimization procedure is dependent on the metric that is
chosen for optimization. We employed the MCC because it includes the
four values from a confusion matrix. Other algorithms such as the
furthest neighbor and nearest neighbor algorithms minimize the number of
FP and FN, respectively; however, these suffer because the number of FN
and FP are not controlled, respectively (13, 16). Alternatively, one
could optimize based on the sensitivity, specificity, or accuracy, which
are each based on two values from the confusion matrix or they could
optimize based on the F1-score, which is based on three values from the
confusion matrix. Because these metrics do not balance all four
parameters equally, it is likely that one parameter will dominate in the
optimization procedure. For example, optimizing for sensitivity could
lead to a large number of FPs. More FPs increases the number of OTUs
while more FNs collapses OTUs together. It is difficult to know which is
worse since community richness and diversity are linked to the number of
OTUs. In addition, increasing the number of FNs would overstate the
differences between communities while increasing the number of FPs would
overstate their similarity. Therefore, it is important to jointly
minimize the number of FPs and FNs. With this in mind, we decided to
optimize utilizing the MCC. It is possible that other metrics that
balance the four parameters could be developed and employed for
optimization of the clustering.

@@

\textbf{\emph{Sequence data and processing steps.}} To evaluate the
OptiClust and the other algorithms we created two synthetic sequence
collections and four sequence collections generated from previously
published studies. The V4 region of the 16S rRNA gene was used from all
datasets because it is a popular region that can be fully sequenced with
two-fold coverage using the commonly used MiSeq sequencer from Illumina
(7). The method for generating the simulated datasets followed the
approach used by Kopylova et al. (34) and Schloss (36). Briefly, we
randomly selected 10,000 uniques V4 fragments from 16S rRNA gene
sequences that were unique from the SILVA non-redundant database (38). A
community with an even relative abundance profile was generated by
specifying that each sequence had a frequency of 100 reads. A community
with a staggered relative abundance profile was generated by specifying
that the abundance of each sequence was a randomly drawn integer sampled
from a uniform distribution between 1 and 200. Sequence collections
collected from human feces (39), murine feces (40), soil (41), and
seawater (42) were used to characterize the algorithms' performance with
natural communities. These sequence collections were all generated using
paired 150 or 250 nt reads of the V4 region. We re-processed all of the
reads using a common analysis pipeline that included quality score-based
error correction (7), alignment against a SILVA reference database (38,
43), screening for chimeras using UCHIME (9), and classification using a
naive Bayesian classifier with the RDP training set requiring an 80\%
confidence score (10).
